Adaptive Sequential Model Selection

### 2015-06-19 notes/nonparam.tex
included final sentence describing sampling scheme in a little more detail

@@

>>>>

{\bf Not sure this ``natural'' analog is the correct one.
That is, it didn't use anything like the linear
function $L$ in my abstract case. Hmm.....}

@@

The sampling scheme now draws samples from $\hat{G}^{*,m}_n$ retaining those 
for which the selected model is the same. With these samples, one can standardize each entry of $\gamma_A(\hat{G}^{*,m}_n)$ by its bootstrap SD and finally compute the observed maxT.

{\bf Again, not sure if this is the ``right'' thing to do as it didn't
use the covariance structure of $yX$ under $\hat{F}_n$ to construct the test.}

### 2015-06-19 paper.tex

More generically, we observe data $Y \in \cY$, with unknown sampling distribution $F$. We then use $Y$ to generate an adaptive sequence of $d$ nested models
\[
M_0(Y) \sub M_1(Y) \sub \cdots \sub M_d(Y) \sub M_\infty.
\]
By ``model,'' we mean a family of probability distributions for $Y$. Without loss of generality, we can take $M_\infty$ as the union of all models under consideration.

@@

\subsubsection{Sparse Parametric Models}

@@

\WFcomment{Shorten this section, move most of it to Discussion}

@@

In our problem the null hypotheses at each step are random and thus so is the completion index $k_0(Y)$. We will thus require {\em conditional} independence of null $p$-values given $k_0$:
\begin{equation}\label{eq:indepCond_random_k0}
  \P(p_{k+1} \leq \alpha_{k+1}, \ldots, p_d \leq \alpha_d
  \mid p_1, \ldots, p_{k_0}, \; k_0 = k) = \prod_{i=k+1}^d \alpha_i
\end{equation}

@@

Sections~\ref{sec:pValsIndep}--\ref{sec:modelSSP} discuss conditions on the model sequence $M_{[d]}$ and the $p$-value sequence $p_{[d]}$ under which~\eqref{eq:stochLargeCond_random_k0} is satisfied.

@@

Consider any filtration
\[
\sF_0 \sub \sF_1 \sub \cdots \sub \sF_d
\]
for which $M_{[k]}$ is measurable with respect to $\sF_k$. We say that $\sF_{[d]}$ {\em separates} the $p$-values $p_{[d]}$ if 
\begin{enumerate}
\item $p_k(Y)$ is conservative given $\sF_{k-1}$ 
  when $F\in M_{k-1}(Y)$, and
\item $p_k(Y)$ is measurable with respect to $\sF_k$.
\end{enumerate}
If $p_k$ is exact, the first requirement is equivalent to requiring that $p_k$ is independent of $\sF_{k-1}$ on the event $F\in M_{k-1}(Y)$.

If we think of $\sF_k$ as representing information available at step $k$, then the first condition means that $p_k$ ``excludes'' whatever evidence may have accrued against the null by step $k-1$, and the second means that any information revealed after step $k$ plays no role in determining $p_k$. In that sense, $\sF_k$ forms a ``wall of separation'' between the $p$-values up to $k$ and the ones after $k$. As a result, separated $p$-values are independent on nulls.

@@

In this section we will consider the problem of constructing a valid selective $p$-value for a single step in the path. At step $k$, we construct $p_k(Y)$ to test
\[
  H_{0,k}:\; F \in M_{k-1}(Y)
  \quad \text{ vs. } \quad
  H_{1,k}:\; F \in M_\infty \setminus M_{k-1}(Y).
\]
As usual, the main complication arises from the fact that the null and alternative hypotheses are random. To simplify matters, we first consider deriving $p$-values for a particular fixed candidate model $M\sub M_{\infty}$.

The random variable  $p_{k,M}(Y)$ is a valid {\em selective $p$-value} for $M$ at step $k$ if it is stochastically larger than uniform under sampling distributions $F\in M$, given that $M$ is selected. That is,
\[
\P_F\left(p_{k,M}(Y) \leq \alpha \mid M_{k-1}(Y) = M\right) 
\leq \alpha, \quad \forall F\in M, \alpha \in [0,1].
\]

Once we have constructed selective $p$-values for each fixed candidate $M$, we can use them as building blocks to construct a combined $p$-value for the random null $M_{k-1}(Y)$. Define
\[
p_k(y) = p_{k, M_{k-1}(y)}(y),
\]
which is a valid $p$-value on the event $\{F \in M_{k-1}(Y)\}$:
\begin{equation}\label{eq:selectiveGuaranteePk}
\P_F\left(p_k \leq \alpha \mid M_{k-1}, 
  \;F\in M_{k-1}\right) \leq \alpha, \quad \forall \alpha \in [0,1].
\end{equation}
Recall that in~\eqref{eq:selectiveGuaranteePk} $M_{k-1}$ is random and not $F$.

In principle, we can construct selective tests for exponential family models conditional on any selection event. See \citet{fithian2014optimal} for a general treatment.

@@

In linear regression, there are several tests that we might perform at step $k$. Let $E_{k-1}$ and $E_k$ denote the active sets for models $M_{k-1}$ and $M_k$ respectively. For simplicity we will assume that only one variable is added at step $k$, so that $E_k = E_{k-1} \cup j_k$.

@@

Many of the selective tests for linear regression use saturated-model tests. See for example \citet{lockhart2014significance}, \citet{taylor2013tests}, \citet{taylor2014exact}, \citet{lee2013exact}, and \citet{loftus2014significance}.

The distinction between the incremental null and the complete null is closely related to the distinction between {\em saturated-model} tests and {\em selected-model} tests discussed in \citet{fithian2014optimal}. Define the saturated model as
\[
M_{\text{sat}}:\; Y \sim \cN(\mu, \sigma^2I_n),
\]
so named because there is a mean parameter for every observation. Because the parameter $\mu$ has the same dimension as the data $Y$, we must assume that $\sigma^2$ is known.

Saturated-model tests assume only that $M_{\text{sat}}$ is true, and perform inference on linear functionals of $\mu$. For example, in linear regression after selecting active set $E$, we might construct confidence intervals for the coefficients $\theta^E$.

In the context of sequential model selection, we can use the saturated-model framework to test the incremental null $H^{\text{inc}}:\; \theta_{j_k}^{E_k} = 0$. The saturated model has $n$-dimensional sufficient statistic $Y$ and natural parameter $\mu$. Letting $\eta_k\in \R^n$ denote the vector for which $\theta_{j_k}^{E_k} = \eta_k'\mu$, the UMPU saturated-model test for $\theta_{j_k}^{E_k}$ is based on the distribution
\begin{equation}\label{eq:satModel}
\L\left( \eta_k'Y \mid E_{k-1}, \;j_k, \;\proj_{\eta_k}^\perp Y \right),
\end{equation}
or equivalently on
\begin{equation}\label{eq:satModelEquiv}
\L\left( X_{j_k}'Y \mid E_{k-1}, \;j_k, \;\proj_{\eta_k}^\perp Y \right).
\end{equation}
The pair $(E_{k-1}, j_k)$ defines the selection event and $\proj_{\eta_k}^\perp Y$ eliminates the nuisance parameter $\proj_{\eta_k}^\perp \mu$.

\subsubsection{Selected-Model Tests}
Selected-model inference~\citep{fithian2014optimal} 
represents a more powerful option when testing for inclusion of variable $j_k$. Selected-model tests are based on the statistical model chosen by our selection procedure. At step $k$, we test the selected null model $M(E_{k-1})$ against either the next model $M(E_k)$ or the upper model $M_\infty$.

One option for selected-model inference is to condition on the identity of $j_k$, the next variable added, and perform the UMPU selective test of $M_{k-1}$ against $M_{k}\setminus M_{k-1}$. If $\sigma^2$ is known, the test is based on
\begin{equation}\label{eq:selModel_cond}
\L\left(X_{j_k}'Y \mid E_{k-1}, \; j_k, \; X_{E_{k-1}}'Y\right).
\end{equation}
Note the contrast between~\eqref{eq:selModel_cond} and~\eqref{eq:satModel}. In~\eqref{eq:selModel_cond}, we only need to condition on an $|E_{k-1}|$-dimensional projection of $Y$, whereas in~\eqref{eq:satModel} we condition on $n-1$-dimensional projection of $Y$. Conditioning on more information tends to sap the power of a test, and we will see in Sections~\ref{sec:bivariate} and~\ref{sec:sparseReg}, the saturated-model tests can pay a heavy price for this extra conditioning.

Instead of conditioning on the next model $M_k$, we could instead use
\begin{equation}\label{eq:selModel_marg}
\L\left(X_{j_k}'Y \mid E_{k-1}, \; X_{E_{k-1}}'Y\right),
\end{equation}
or even replace $X_{j_k}'Y$ with any other test statistic $T(Y)$. Under the null hypothesis, the distribution of $Y$ is completely known once we condition on the value of $X_{E_{k-1}}'Y$, the complete sufficient statistic of $M_{k-1}$.

@@

For some applications, a major conceptual advantage of the saturated-model approach is that, when $M(E)$ is misspecified, $\theta^{E}$ is still a well-defined quantity on which we can perform exact inference. For example, a 95\% confidence interval for $\theta_{j_k}^{E_k}$ would cover that coefficient with probability 95\% even if $M(E_k)$ is misspecified. In the setting of sequential model selection, this advantage is less compelling. We do not necessarily need a scientifically interpretable interval for $\theta_{j_k}^{E_k}$; rather, the goal of the test at step $k$ is to reject and move on if $M(E_{k-1})$ is demonstrably incomplete.

@@

Second, we will see in Section~\ref{sec:pValsIndep} that tests of the form~\eqref{eq:selModel_cond} or~\eqref{eq:selModel_marg} yield independent $p$-values under general conditions, allowing us to apply  the sequential stopping rules of~\citet{gsell2013sequential} and \citet{li2015accumulation}. Finally, and perhaps most importantly, selected-model tests can be dramatically more powerful than saturated-model tests, as we will see next.

In this section we illustrate the superior power of the selected-model test in early steps with an extended example.

>>>>

More generically, we observe data $Y \in \cY$, with unknown sampling distribution $F$. We then use $Y$ to generate an adaptive sequence of $d$ nested models
\[
M_0(Y) \sub M_1(Y) \sub \cdots \sub M_d(Y) \sub M_\infty.
\]
\WFcomment{should probably end with $M_{d-1}(Y)$. need to fix this   throughout paper.}
By ``model,'' we mean a family of probability distributions for $Y$. Without loss of generality, we can take $M_\infty$ as the union of all models under consideration.

@@

\subsubsection{Sparse Parametric Models}\label{sec:genSparse}

@@

\WFcomment{Shorten this section, move some of it to Discussion?}

@@

In our problem the null hypotheses at each step are random and thus so is the completion index $k_0(Y)$. We will thus require {\em conditional} independence of null $p$-values given $k_0$:
\begin{equation}\label{eq:indepCond_random_k0}
  \P(p_{k+1} \leq \alpha_{k+1}, \ldots, p_d \leq \alpha_d
  \mid p_1, \ldots, p_k, \; k_0 = k) = \prod_{i=k+1}^d \alpha_i
\end{equation}

@@

Sections~\ref{sec:pValsIndep}--\ref{sec:SSP} discuss conditions on the model sequence $M_{[d]}$ and the $p$-value sequence $p_{[d]}$ under which~\eqref{eq:stochLargeCond_random_k0} is satisfied.

@@

Consider any filtration
\[
\sF_0 \sub \sF_1 \sub \cdots \sub \sF_d
\]
for which $M_{[k]}$ is measurable with respect to $\sF_k$. We say that $\sF_{[d]}$ {\em separates} the $p$-values $p_{[d]}$ if 
\begin{enumerate}
\item $p_k(Y)$ is conservative given $\sF_{k-1}$ 
  when $F\in M_{k-1}(Y)$, and
\item $p_k(Y)$ is measurable with respect to $\sF_k$.
\end{enumerate}
If $p_k$ is exact, the first requirement is equivalent to requiring that $p_k$ is independent of $\sF_{k-1}$ on the event $F\in M_{k-1}(Y)$ (i.e., the event when $p_k$ is null).

If we think of $\sF_k$ as representing information available at step $k$, then the first condition means that $p_k$ ``excludes'' whatever evidence may have accrued against the null by step $k-1$, and the second means that any information revealed after step $k$ plays no role in determining $p_k$. In that sense, $\sF_k$ forms a ``wall of separation'' between the $p$-values up to $k$ and the null $p$-values after $k$. As a result, separated $p$-values are independent on nulls.

@@

Having derived conditions for independence, we turn now to the question of how to construct valid selective $p$-values for a single step in the path. At step $k$, we construct $p_k(Y)$ to test
\[
  H_{k}:\; F \in M_{k-1}(Y)
\]
against the alternative $F \in M_\infty \setminus M_{k-1}(Y)$. As usual, the main complication arises from the fact that the null hypothesis is random. To simplify matters, we first consider deriving $p$-values for a particular fixed candidate model $M\sub M_{\infty}$.

The random variable  $p_{k,M}(Y)$ is a valid {\em selective $p$-value} for $M$ at step $k$ if it is conservative under sampling from $F\in M$, given that $M$ is selected. That is,
\[
\P_F\left(p_{k,M}(Y) \leq \alpha \mid M_{k-1}(Y) = M\right) 
\leq \alpha, \quad \forall F\in M, \alpha \in [0,1].
\]

Once we have constructed selective $p$-values for each fixed candidate $M$, we can use them as building blocks to construct a combined $p$-value for the random null $M_{k-1}(Y)$. Define
\[
p_k(y) = p_{k, M_{k-1}(y)}(y),
\]
which is a valid $p$-value on the event $\{F \in M_{k-1}(Y)\}$:
\begin{equation}\label{eq:selectiveGuaranteePk}
\P_F\left(p_k \leq \alpha \mid M_{k-1}, 
  \;F\in M_{k-1}\right) \leq \alpha, \quad \forall \alpha \in [0,1].
\end{equation}
%Recall that in~\eqref{eq:selectiveGuaranteePk} $M_{k-1}$ is random and not $F$.

One very useful strategy for constructing exact selective tests is to condition on the sufficient statistics of the null model. Because the law
\[
\L(Y \;\mid\; T_{k-1}, \; M_{k-1}=M)
\]
is known under any $F\in M$, we can construct selective tests by comparing the distribution of any test statistic $U(Y)$ to its known conditional distribution under the null. See \citet{fithian2014optimal} for a general treatment of selective tests in exponential family models.

If we follow this strategy, then the reference distribution for $U(Y)$ is measurable with respect to $T_{k-1}$ and $M_{k-1}$; moreover, if $U(Y)$ is also measurable with respect to $(M_k, T_k)$, the next model in the path and its sufficient statistics, then $p_k$ is measurable with respect to $\sF_k$, and so the $p$-value sequence $p_{[d]}$ will be independent on nulls per Theorem~\ref{thm:suffCond}.

For the remainder of this section we will focus on linear regression.

@@

In linear regression, there are several types of selective test that we might contemplate performing at step $k$. Let $E_{k-1}$ and $E_k$ denote the active sets for models $M_{k-1}$ and $M_k$ respectively. For simplicity we will assume that only one variable is added at step $k$, so that $E_k = E_{k-1} \cup j_k$. It is relatively straightforward to generalize the developments in this section to the case where multiple variables are added at once.

@@

Many of the selective tests for linear regression use the saturated-model framework. See for example \citet{lockhart2014significance}, \citet{taylor2013tests}, \citet{taylor2014exact}, \citet{lee2013exact}, and \citet{loftus2014significance}.

The distinction between the incremental null and the complete null is closely related to the distinction between {\em saturated-model} tests and {\em selected-model} tests. Define the saturated model as
\[
M_{\text{sat}}:\; Y \sim \cN(\mu, \sigma^2I_n),
\]
so named because there is a mean parameter for every observation. Because the parameter $\mu$ has the same dimension as the data $Y$, we must assume that $\sigma^2$ is known.

Saturated-model tests assume only that $M_{\text{sat}}$ is true, and perform inference on linear functionals of $\mu$. For example, in linear regression after selecting active set $E$, we might construct confidence intervals for the least-squares coefficients $\theta^E$ of active set $E$.

In the context of sequential model selection, we can use the saturated-model framework to test the incremental null $H^{\text{inc}}:\; \theta_{j_k}^{E_k} = 0$. The saturated model has $n$-dimensional sufficient statistic $Y$ and natural parameter $\mu$. Letting $\eta_k\in \R^n$ denote the vector for which $\theta_{j_k}^{E_k} = \eta_k'\mu$, the UMPU saturated-model test for $\eta_k'\mu$ is based on the distribution
\begin{equation}\label{eq:satModel}
\L\left( \eta_k'Y \mid E_{k-1}, \;j_k, \;\proj_{\eta_k}^\perp Y \right),
\end{equation}
or equivalently on
\begin{equation}\label{eq:satModelEquiv}
\L\left( X_{j_k}'Y \mid E_{k-1}, \;j_k, \;\proj_{\eta_k}^\perp Y \right).
\end{equation}
The pair $(E_{k-1}, j_k)$ defines the selection event and $\proj_{\eta_k}^\perp Y$ eliminates the nuisance parameter $\proj_{\eta_k}^\perp \mu$.

While the test statistic $\eta_k'Y$ is measurable with respect to $\sF_k$, the reference distribution we compare it against is not. Specifically, $\proj_{\eta_k}^\perp Y$ is not measurable with respect to $\sF_k$. As a result, saturated-model $p$-values constructed as in~\eqref{eq:satModel} are in general {\em not} independent on nulls.

\subsubsection{Selected-Model Tests}
Selected-model inference~\citep{fithian2014optimal} 
represents a more powerful option when testing for inclusion of variable $j_k$. While saturated-model inference always uses the same model $M_{\text{sat}}$, selected-model inference is based on the statistical model that is chosen by our selection procedure. In the sequential case this means that at step $k$, we test the selected null model $M(E_{k-1})$, against either the next selected model $M(E_k)$ or the upper model $M_\infty$.

\paragraph{Conditioning on the Next Model}
One option for selected-model inference is to condition on the identity of $j_k$, the next variable added, and perform a UMPU, equal-tailed, or one-sided selective test of $M_{k-1}$ against $M_{k}\setminus M_{k-1}$ conditional on both models. If $\sigma^2$ is known, the test is based on
\begin{equation}\label{eq:selModel_cond}
\L\left(X_{j_k}'Y \mid E_{k-1}, \; j_k, \; X_{E_{k-1}}'Y\right).
\end{equation}
Note the contrast between~\eqref{eq:selModel_cond} and~\eqref{eq:satModel}. In~\eqref{eq:selModel_cond}, we only need to condition on an $|E_{k-1}|$-dimensional projection of $Y$, whereas in~\eqref{eq:satModel} we condition on $n-1$-dimensional projection of $Y$. Conditioning on more information tends to sap the power of a test, and as we will see in Sections~\ref{sec:bivariate} and~\ref{sec:sparseReg}, the saturated-model tests can pay a heavy price for this extra conditioning.

If we use the construction in~\eqref{eq:selModel_cond} after a subpath-sufficient path algorithm, the $p$-values are always guaranteed to have independence on nulls. The test statistic $X_{j_k}'Y$ is measurable with respect to $\sF_k$, and so is the reference distribution, which depends on $E_{k-1}, j_k,$ and $T_{k-1} = X_{E_{k-1}}'Y$.

\paragraph{The Max-$t$ and Next-$\lambda$ Tests}
Instead of conditioning on the next model $M_k$, we could use
\begin{equation}\label{eq:selModel_marg}
\L\left(X_{j_k}'Y \mid E_{k-1}, \; X_{E_{k-1}}'Y\right),
\end{equation}
or even replace $X_{j_k}'Y$ with any other test statistic $U(Y)$. As long as $U(Y)$ is measurable with respect to $(M_k, T_k)$, and the path algorithm is subpath-sufficient, the $p$-values are independent on nulls.

Two specific examples are worth discussing in detail. First, consider the {\em selective max-$t$ test}, a generalization of the max-$t$ test of ~\citet{bujaxxxxmaxT}. Controlling for $E$, the $z$-statistic for coefficient $j\notin E$ is
\begin{equation}
  z_{j,E}(Y) = \frac{(\proj_E^\perp X_j)'Y}
  {\sigma\|\proj_{E}^\perp X_j\|_2},
\end{equation}
and the $t$-statistic is
\begin{equation}
  t_{j,E}(Y) = \frac{(\proj_E^\perp X_j)'Y}
  {\hat \sigma_{j,E} \|\proj_{E}^\perp X_j\|_2},
  \qquad \text{where} \quad 
  \hat \sigma_{j,E}^2(Y) = 
  \frac{\|\proj_{E \cup \{j\}}^\perp Y\|_2^2}{n-|E|-1}.
\end{equation}
Note that $z_{j,E}$ is a monotonic function of $X_j'Y$ given $X_E'Y$, and $t_{j, E}$ is a monotonic function of $X_j'Y$ given $(X_E'Y, \|Y\|_2)$. Thus $z_{j,E}$ is measurable with respect 
to $X_{E \cup \{j\}}'Y$ if $\sigma^2$ is known, and otherwise $t_{j,E}$ is measurable with respect to $(X_{E \cup \{j\}}'Y, \|Y\|_2)$.

We can then define the max-$t$-statistic (respectively max-$z$-statistic) for active set $E$ as
\begin{equation}\label{eq:maxT}
  U_E(Y) = \max_{j \notin E} \left|t_{j,E}(Y)\right|,
\end{equation}
and use $U_{E_{k-1}}(Y)$ as our test statistic at step $k$.

$U_{E_{k-1}}(Y)$ is $\sF_k$-measurable if and only if $j_k$ is the maximizer in~\eqref{eq:maxT}. Consequently, max-$t$ $p$-values are independent on nulls if they are used in tandem with forward-stepwise regression, where the next variable added is always the variable with the largest $t$-statistic.

By contrast, the LASSO path does not in general add the variable with the largest $t$-statistic. In that case, we could use instead the {\em selective next-$\lambda$ test}, which uses test statistic
\begin{equation}
  U_k = \inf\{ \lambda:\; \hat\beta^\lambda \notin \beta(E_{k-1})\}.
\end{equation}
The next-$\lambda$ statistic is $\sF_k$-measurable if we use the LASSO to select variables, but not if we use forward-stepwise. \WFcomment{Relate this to spacings test. Is this essentially the ``generalized spacings test?''}

@@

For some applications, a major conceptual advantage of the saturated-model approach is that, when $M(E)$ is misspecified, $\theta^{E}$ is still a well-defined quantity on which we can perform exact inference. For example, a 95\% confidence interval for $\theta_{j_k}^{E_k}$ could cover that coefficient with probability 95\% even if $M(E_k)$ is misspecified. In the setting of sequential model selection, this advantage is less important. We do not necessarily need a scientifically interpretable interval for $\theta_{j_k}^{E_k}$; rather, the goal of the test at step $k$ is to reject $M_{k-1}$ and move on if it is demonstrably incompatible with the data.

@@

Second, we have seen in Section~\ref{sec:pValsIndep} that tests of the form~\eqref{eq:selModel_cond} or~\eqref{eq:selModel_marg} yield independent $p$-values under general conditions, allowing us to apply  the sequential stopping rules of~\citet{gsell2013sequential} and \citet{li2015accumulation}.

Finally, and perhaps most importantly, selected-model tests can be dramatically more powerful than saturated-model tests, as we illustrate now with an extended example.
